Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-28 paper/paper.tex
latest - 28th

\warning{The  semantics  of   expressions  (and  applications)
%% \footnote{In  most
%%   contexts, we  shall use the  term 'expression' and the  notation $e$
%%   for both expressions and applications.} 
are given by transitions of the form $\rho, \stk, \heap, e \rightarrow
\rho',  \stk', \heap',  e'$.  Here  \stk\ is  a stack  of continuation
frames.  Each  continuation frame  is of the  form $(\rho,  \ell, e)$,
signifying that the  location $\ell$ has to be updated  with the value
of the  currently evaluating  expression, and $e$  is to  be evaluated
next  in the  environment $\rho$.   We shall call the $\rho$ and
the $e$ on the LHS of a $ \stk,
\heap,  e$ is the state of  the  transition system, then we shall there is a   \emph{current
  evaluation  context}, given  by  $\rho$, $\heap$  and  $e$, and  
\emph{suspended  evaluation  contexts}  (represented  by the
environment, heap and expression in each   continuation
frame) in the stack $S$.}
%% \warning{The
%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
%% system will be denoted $\mathcal{S}$.}

The  initial  state  of  the  transition  system  described  above  is
$([\;]_\rho,\,   (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
(\print~\ans)):[\;]_{S}, 
[\;]_{\heap},\, (\mainpgm))$,  in which  $[\;]_\rho$,
$[\;]_{\mbox \footnotesize {H}}$
and $[\;]_{S}$ are    are the
empty environment, heap and stack respectively. The  initial stack consists
of  a single  continuation frame  in  which \ans\  is a  distinguished
variable  that  would   eventually  be  updated  with   the  value  of
(\mainpgm) and $\rho_\mathit{init}$ maps \ans\ to a location $\ell_{\ans}$.   In  addition, \print\ is  a function modeling  a printing
mechanism---a   standard   run-time   support  assumption   for   lazy
languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The operator $:$
pushes elements on top of the stack.

@@

A notion  that generalizes liveness  is {\em demand}.   While liveness
gives the patterns of future uses of a variable, demand represents the
future use of the value of an expression.  The demand on an expression
$e$   is   again   a   set    of   access   paths---that   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the  let expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   \SRETURN~x)$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the  set  of $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.

@@

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
replacing   $\nont{\mathit{all}}{}$   in   the  sentential   form   by
$\epsilon$. However, if the TM does  not accept the input string, then
any  sentential  form derived  from  the  start  symbol would  have  a
non-terminal different from $\nont{\mathit{final}}{c}$. Thus the empty
string would not be derivable from the grammar.
\end{proof}

>>>>

The  semantics  of   expressions  (and  applications\footnote{In  most
 contexts, we  shall use the  term 'expression' and the  notation $e$
  for both expressions and applications.}) 
are given by transitions of the form $\rho, \stk, \heap, e \rightsquigarrow
\rho',  \stk', \heap',  e'$.  Here  \stk\ is  a stack  of continuation
frames.  Each  continuation frame  is of the  form $(\rho,  \ell, e)$,
signifying that the  location $\ell$ has to be updated  with the value
of the  currently evaluating  expression, and $e$  is to  be evaluated
next  in the  environment $\rho$.  
%% Given a state $\rho, \stk, \heap,
%% e$  of  the  transition system, we shall call the pair $(\rho, e)$ as
%% the    \emph{current
%%   evaluation  context}, and for a continuation frame $(\rho',  \ell, e')$ in
%% \stk, we shall call the pair $(\rho', e')$ as a   
%% \emph{suspended  evaluation  context}.}
%% %% \warning{The
%% 3-tuple $\rho, \stk,  \heap, e$ representing the state  of the transition
%% system will be denoted $\mathcal{S}$.}
The  initial  state  of  the  transition  system  described  above  is
$([\;]_\rho,\,   (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
(\print~\ans)):[\;]_{S}, 
[\;]_{\heap},\, (\mainpgm))$,  in which  $[\;]_\rho$,
$[\;]_{\mbox \footnotesize {H}}$
and $[\;]_{S}$ are    are the
empty environment, heap and stack respectively. The  initial stack consists
of  a single  continuation frame  in  which \ans\  is a  distinguished
variable  that  would   eventually  be  updated  with   the  value  of
(\mainpgm) and $\rho_\mathit{init}$ maps \ans\ to a location $\ell_{\ans}$.   In  addition, \print\ is  a function modeling  a printing
mechanism---a   standard   run-time   support  assumption   for   lazy
languages~\cite{Jones87}---that prints  the value of  (\mainpgm).  The operator $:$
pushes elements on top of the stack.

@@

A notion  that generalizes liveness  is {\em demand}.   While liveness
gives the patterns of future uses of a variable, demand represents the
future uses of the value of an expression.  The demand on an expression
$e$   is   again   a   set    of   access   paths---that   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the  let expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   \SRETURN~x)$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the  set  of $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.

@@

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
replacing   $\nont{\mathit{all}}{}$   in   the  sentential   form   by
$\epsilon$. Conversely,  if the TM  does not accept the  input string,
then  any sentential  form derived  from the  start symbol  would have
exactly one non-terminal, and this non-terminal will be different from
$\nont{\mathit{final}}{c}$.  Thus  the  empty   string  would  not  be
derivable from the grammar.
\end{proof}
